Labor Markets and Technological Change: Evidence from Electronic Health Records

### 2022-03-18 Paper/Paper.tex

The relationship between technological innovation and labor markets is a longstanding topic of interest in economics and in policy making. Their interaction has direct impacts across industries; job changes such as displacement, creation, and satisfaction are all prevalent for many workers (\cite{autor2003skill}, \cite{fallick1996review}, \cite{akerlof1988job}). Labor market changes for one job type may also have rippling effects on populations serviced by that industry. One such job is that of a physician, where physician choices and labor markets directly affect patients on margins of accessibility and quality of care received. For example, physician burnout is correlated with lower patient satisfaction (\cite{shanafelt2002burnout}), physician shortages are linked to higher mortality rates (\cite{gong2019higher}), and administrative regulations such as standardized checklists may be associated with lower mortality in surgical settings (\cite{treadwell2014surgical}). Thus, shocks to physician labor markets or work settings may have drastic implications for patients. I investigate physician labor markets under a major technology change in the U.S. healthcare system, the implementation of electronic health care record systems, which have been emphasized as a cause of burnout and may induce changes to physician labor market decisions.

Electronic Health Records (EHRs) are computerized medical records which include detailed accounts and notes of medical history, stored in advanced systems with additional capabilities such as providing suggesions for care. EHRs have have become increasingly relevant in the U.S. since 2008, in part after former President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}). The movement towards digitization in health care was expected to have immediate and substantial impacts on efficiency. A 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}). Such expectations led the U.S. government to incentivize the use of EHRs in hospitals with the the Health Information Technology for Economic and Clinical Health (HITECH) Act in 2008 (\cite{hitech})\footnote{This legislation subsidized hospitals which used EHRs “meaningfully”According to Quatris Healthco, meaningful use standards proceeded in three stages over time. In Stage 1 (2010), MU focused on data capturing and sharing. In Stage 2, which began in late 2012, MU extended to using EHRs for patient incorporation and using the technology as a helper in care. Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})}. The percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}), evidence of the nationwide movement towards digitization.

Physicians, as primary users of this technology, have a meaningful role in whether potential benefits of EHRs are realized. However, with the rapid implementation of EHRs, the day to day life of physicians also changed rapidly. In interviews, physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels. Senior physicians in particular were found to “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}). The frustration of using a new technology raises the cost of working in specific hospitals, which may lead physicians on the margin to make changes in labor market choices such as exiting the labor force altogether. Further, the short and long run impacts of EHRs on physician productivity could vary greatly due to the immediate learning curve that subsides over time.  I use difference in difference with heterogeneous treatment timing to understand the effect of EHR implementation in hospitals on physician labor market outcomes: retirement, choice of work setting, and patients seen in the short vs. long run.

@@

\section{Background}

@@

The HITECH Act was passed in 2009, setting aside government funds to subsidize hospitals who used EHRs according to certain guidelines. These requirements include having at least 80\% of patients in the system, recording answers to specific questions, and protecting the security of the system to ensure privacy. Immediately, EHR companies took advantage of the legislation and began marketing campaigns centered around their system being eligible to earn the hospital subsidies. Hospitals began receiving payments for the first stage of meaningful use in 2011. This led to a 75 percentage point increase in the number of hospitals with EHRs from 2009 to 2015 (\cite{stats}). \textcolor{red}{maybe put HealthIT US map that shows hospitals who received incentive payment?} While EHRs are widely implemented in hospitals by 2015, physicians still continue to report frustration. "Some healthcare professionals continue to develop workarounds and rely on paper alternatives rather than using EHRs" (\cite{evans2016electronic}).

@@

I include means of these variables in physicians either younger than or older than 45 in \hyperref[tab:sumstats2]{Table \ref{tab:sumstats2}}. Senior physicians see more patients and work in more hospitals on average than younger physicians. Physicians over 45 are slightly more likely to be exposed to EHRs in both general hospitals and low integration hospitals, but a smaller fraction of their hospitals utilize EHRs. These older physicians also see a larger fraction of patients in an office setting and are more likely to work in an office in general. Finally, physicians in the higher age bracket are less likely to switch zip codes. \textcolor{red}{why is there equal retirement between young and old?}

@@

Figure \ref{fig:retirefirst} shows the aggregated group time treatment effects of being exposed to an EHR in any hospital. The left panel of the figure includes all physicians and shows a clear increase in the probability of retiring after exposure to an EHR. On the right panel, only physicians over the age of 45 in 2017 are included.

\begin{figure}[ht]
\caption{Event Study Graphs}
\vspace{2mm}
        \begin{minipage}[b]{0.47\linewidth}
            \centering
            \includegraphics[width=\textwidth]{Objects/ggdid_retire_allEHR.pdf}
        \end{minipage}
        \hspace{0.2cm}
        \begin{minipage}[b]{0.47\linewidth}
            \centering
            \includegraphics[width=\textwidth]{Objects/ggdid_retire_allEHR_old.pdf}
        \end{minipage}
        \label{fig:retirefirst}
\end{figure}

>>>>

The relationship between technological innovation and labor markets is a longstanding topic of interest in economics and in policy making. Their interaction has a direct impact across industries; job characteristics such as displacement, creation, and satisfaction are all prevalent for many workers (\cite{autor2003skill}, \cite{fallick1996review}, \cite{akerlof1988job}). Labor market changes for one job type may have rippling effects on populations serviced by that industry. One such job is that of a physician, where physician choices and labor markets directly affect patient accessibility and quality of care received. For example, physician burnout is correlated with lower patient satisfaction (\cite{shanafelt2002burnout}), physician shortages are linked to higher mortality rates (\cite{gong2019higher}), and administrative regulations on physicians such as standardized checklists may be associated with lower mortality in surgical settings (\cite{treadwell2014surgical}). Thus, shocks to physician labor markets or work settings potentially have drastic implications for patients. I investigate physician labor market changes as a result of a major technology change in the U.S. healthcare system, the implementation of electronic health care record systems, which have been emphasized as a cause of burnout and may induce changes to physician labor market decisions.

Electronic Health Records (EHRs) are computerized medical records which include detailed accounts and notes of medical history, stored in advanced systems with additional capabilities such as providing suggestions for care. EHRs have have become increasingly relevant in the U.S. since 2008, in part after former President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}). The movement towards digitization in health care was expected to have immediate and substantial impacts on efficiency. A 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}). Such expectations led the U.S. government to incentivize the use of EHRs in hospitals with the the Health Information Technology for Economic and Clinical Health (HITECH) Act in 2008 (\cite{hitech})\footnote{This legislation subsidized hospitals which used EHRs “meaningfully”According to Quatris Healthco, meaningful use standards proceeded in three stages over time. In Stage 1 (2010), MU focused on data capturing and sharing. In Stage 2, which began in late 2012, MU extended to using EHRs for patient incorporation and using the technology as a helper in care. Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})}. The percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}), revealing the nationwide movement towards digitization.

Physicians, as primary users of this technology, play a meaningful role in whether potential benefits of EHRs are realized. However, with the rapid implementation of EHRs, the day to day life for physicians changed rapidly. In interviews, physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels. Senior physicians in particular were found to “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}). The frustration of using a new technology raises the cost of working in specific hospitals, which may lead physicians on the margin to make changes in labor market choices such as exiting the labor market altogether. Further, the short and long run impacts of EHRs on physician productivity could vary greatly due to the immediate learning curve that subsides over time.  Treating EHR implementation as an exogenous treatment vairble, I use difference in difference with heterogeneous treatment timing to understand the effect of EHR implementation in hospitals on physician labor market outcomes: retirement, choice of work setting, and amount of patients seen.

@@

\section{Institutional Details}

@@

The HITECH Act was passed in 2009, setting aside government funds to subsidize hospitals who used EHRs according to certain guidelines. These requirements include having at least 80\% of patients in the system, recording answers to specific questions, and protecting the security of the system to ensure privacy. Immediately, EHR companies took advantage of the legislation and began marketing campaigns centered around their system being eligible to earn the hospital subsidies. Hospitals began receiving payments for the first stage of meaningful use in 2011. This led to a 75 percentage point increase in the number of hospitals with EHRs from 2009 to 2015 (\cite{stats}). \textcolor{red}{maybe put HealthIT US map that shows hospitals who received incentive payment?} While EHRs are widely implemented in hospitals by 2015, physicians still continue to report frustration. "Some healthcare professionals continue to develop workarounds and rely on paper alternatives rather than using EHRs" (\cite{evans2016electronic}).  
\textcolor{red}{add details about how this technology impacts physician day to day life}

@@

I include means of these variables in physicians either younger than or older than 60 in \hyperref[tab:sumstats2]{Table \ref{tab:sumstats2}}. Senior physicians see more patients and work in more hospitals on average than younger physicians. These older physicians also see a larger fraction of patients in an office setting and are more likely to work in an office in general. Finally, physicians in the higher age bracket are less likely to switch zip codes. The age groups are exposed to EHRs in similar ways. \textcolor{red}{why is there equal retirement between young and old?}

@@

Figure \ref{fig:retirefirst} shows the aggregated group time treatment effects of being exposed to an EHR in any hospital. The left hand panel includes all physicians and shows no effect of EHR implementation on retirement. When limiting the sample to only physicians 60 and above (retirement age), the year of EHR implementation leads to a .37 percentage point increase in the probability of retiring.

\vspace{5mm}
\begin{figure}[ht]
\caption{Event Study Graphs}
\vspace{2mm}
        \begin{minipage}[b]{0.47\linewidth}
            \centering
            \includegraphics[width=\textwidth]{Objects/ggdid_retire_allEHR.pdf}
        \end{minipage}
        \hspace{0.2cm}
        \begin{minipage}[b]{0.47\linewidth}
            \centering
            \includegraphics[width=\textwidth]{Objects/ggdid_retire_allEHR_old.pdf}
        \end{minipage}
        \label{fig:retirefirst}
\end{figure}
\vspace{5mm}
